---
title: "Tidyverse in Action"
subtitle: "A hands-on demo with ggplot2, dplyr, tidyr & more"
author: "Workshop Demo"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    code-fold: false
    code-tools: true
    highlight-style: github
    fig-width: 9
    fig-height: 5.5
    embed-resources: true
execute:
  warning: false
  message: false
---

# Setup

Load the tidyverse — this brings in **ggplot2**, **dplyr**, **tidyr**, **readr**, **stringr**, **forcats**, and more.

```{r}
library(tidyverse)
```

---

# 1. The Dataset

We build a realistic sales dataset from scratch — no external files needed.

```{r}
set.seed(42)

products   <- c("Laptop", "Tablet", "Phone", "Monitor", "Keyboard")
regions    <- c("North", "South", "East", "West")
months     <- seq(as.Date("2023-01-01"), as.Date("2023-12-01"), by = "month")
base_price <- c(Laptop = 1200, Tablet = 450, Phone = 800, Monitor = 350, Keyboard = 80)

sales_raw <- expand_grid(month = months, product = products, region = regions) |>
  mutate(
    units = rpois(n(), lambda = case_when(
      product == "Keyboard" ~ 60,
      product == "Phone"    ~ 45,
      product == "Tablet"   ~ 30,
      product == "Monitor"  ~ 20,
      TRUE                  ~ 15)),
    units      = as.integer(units * if_else(month(month) %in% 10:12, 1.4, 1)),
    price      = base_price[product] * runif(n(), 0.92, 1.08),
    revenue    = units * price,
    category   = case_when(
      product %in% c("Laptop", "Monitor") ~ "Computing",
      product %in% c("Phone",  "Tablet")  ~ "Mobile",
      TRUE                                ~ "Accessories"),
    margin_pct = case_when(
      category == "Computing"   ~ runif(n(), 0.18, 0.28),
      category == "Mobile"      ~ runif(n(), 0.22, 0.35),
      TRUE                      ~ runif(n(), 0.35, 0.55))
  )

glimpse(sales_raw)
```

---

# 2. Data Wrangling with `dplyr`

## 2.1 Summary by Product

```{r}
product_summary <- sales_raw |>
  group_by(product, category) |>
  summarise(
    total_revenue = sum(revenue),
    total_units   = sum(units),
    avg_price     = mean(price),
    avg_margin    = mean(margin_pct),
    .groups = "drop"
  ) |>
  arrange(desc(total_revenue))

product_summary
```

## 2.2 Monthly Trends

```{r}
monthly <- sales_raw |>
  group_by(month, category) |>
  summarise(revenue = sum(revenue), units = sum(units), .groups = "drop")

monthly |> slice_head(n = 8)
```

## 2.3 Pivoting with `tidyr`

```{r}
sales_raw |>
  group_by(product, region) |>
  summarise(revenue = sum(revenue), .groups = "drop") |>
  pivot_wider(names_from = region, values_from = revenue) |>
  mutate(across(where(is.numeric), \(x) scales::dollar(x, scale = 1e-3, suffix = "k")))
```

---

# 3. Visualisation with `ggplot2`

## 3.1 Revenue by Product — Bar Chart

```{r}
product_summary |>
  mutate(product = fct_reorder(product, total_revenue)) |>
  ggplot(aes(x = total_revenue, y = product, fill = category)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = scales::dollar(total_revenue, scale = 1e-3, suffix = "k")),
            hjust = -0.1, size = 3.5, fontface = "bold") +
  scale_x_continuous(
    labels = scales::dollar_format(scale = 1e-3, suffix = "k"),
    expand = expansion(mult = c(0, 0.18))
  ) +
  scale_fill_manual(values = c(Computing = "#2D6A4F", Mobile = "#52B788",
                               Accessories = "#B7E4C7")) +
  labs(title = "Total Annual Revenue by Product",
       subtitle = "Laptops lead despite lower unit volume",
       x = "Revenue (USD)", y = NULL, fill = "Category") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        panel.grid.major.y = element_blank(),
        legend.position = "bottom")
```

## 3.2 Monthly Revenue Trend — Line Chart

```{r}
monthly |>
  ggplot(aes(x = month, y = revenue, color = category, group = category)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  annotate("rect", xmin = as.Date("2023-10-01"), xmax = as.Date("2023-12-31"),
           ymin = -Inf, ymax = Inf, alpha = 0.08, fill = "#F4A261") +
  annotate("text", x = as.Date("2023-11-01"), y = Inf,
           label = "Q4 Boost", vjust = 2, size = 3.5, color = "#E76F51") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  scale_y_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  scale_color_manual(values = c(Computing = "#2D6A4F", Mobile = "#52B788",
                                Accessories = "#D4A017")) +
  labs(title = "Monthly Revenue by Category — 2023",
       subtitle = "All categories show a Q4 seasonal uplift",
       x = NULL, y = "Revenue (USD)", color = "Category") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```

## 3.3 Units by Region — Faceted Bar Chart

```{r}
#| fig-height: 6

sales_raw |>
  group_by(region, product, category) |>
  summarise(units = sum(units), .groups = "drop") |>
  mutate(product = fct_reorder(product, units)) |>
  ggplot(aes(x = units, y = product, fill = category)) +
  geom_col(width = 0.7) +
  facet_wrap(~region, ncol = 2) +
  scale_x_continuous(labels = scales::comma) +
  scale_fill_manual(values = c(Computing = "#2D6A4F", Mobile = "#52B788",
                               Accessories = "#D4A017")) +
  labs(title = "Units Sold by Region and Product",
       subtitle = "Keyboard dominates unit volume in every region",
       x = "Units Sold", y = NULL, fill = "Category") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 12),
        panel.grid.major.y = element_blank(),
        legend.position = "bottom")
```

## 3.4 Margin vs Revenue — Scatter Plot

```{r}
sales_raw |>
  group_by(product, category, month) |>
  summarise(revenue = sum(revenue), margin = mean(margin_pct),
            units = sum(units), .groups = "drop") |>
  ggplot(aes(x = revenue, y = margin, color = category, size = units)) +
  geom_point(alpha = 0.65) +
  geom_smooth(aes(group = category), method = "lm", se = FALSE,
              linewidth = 0.8, linetype = "dashed") +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_size_continuous(range = c(2, 8), labels = scales::comma) +
  scale_color_manual(values = c(Computing = "#2D6A4F", Mobile = "#52B788",
                                Accessories = "#D4A017")) +
  labs(title = "Margin % vs Monthly Revenue",
       subtitle = "Accessories carry higher margins despite lower revenue",
       x = "Monthly Revenue", y = "Gross Margin %",
       color = "Category", size = "Units Sold") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

## 3.5 Heatmap — Revenue by Product × Month

```{r}
sales_raw |>
  group_by(product, month) |>
  summarise(revenue = sum(revenue), .groups = "drop") |>
  mutate(month_label = fct_reorder(format(month, "%b"), month)) |>
  ggplot(aes(x = month_label, y = fct_reorder(product, revenue), fill = revenue)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = scales::dollar(revenue, scale = 1e-3, suffix = "k", accuracy = 1)),
            size = 3, color = "white", fontface = "bold") +
  scale_fill_gradient(low = "#B7E4C7", high = "#1B4332",
                      labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  labs(title = "Revenue Heatmap — Product x Month",
       subtitle = "Darker = higher revenue",
       x = NULL, y = NULL, fill = "Revenue") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        panel.grid = element_blank())
```

---

# 4. `stringr` & `forcats`

## 4.1 String manipulation

```{r}
sales_raw |>
  distinct(product, category) |>
  mutate(
    label      = str_glue("[{str_pad(category, 12, 'right')}] {str_to_upper(product)}"),
    slug       = str_to_lower(str_replace_all(product, " ", "-")),
    has_mobile = str_detect(category, "Mobile")
  )
```

## 4.2 Factor reordering

```{r}
sales_raw |>
  mutate(product = fct_reorder(product, revenue, .fun = sum)) |>
  group_by(product) |>
  summarise(revenue = sum(revenue)) |>
  ggplot(aes(x = revenue, y = product)) +
  geom_col(fill = "#2D6A4F") +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  labs(title = "fct_reorder() — Products sorted by revenue",
       x = "Total Revenue", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        panel.grid.major.y = element_blank())
```

---

# 5. Summary Table

```{r}
sales_raw |>
  group_by(category) |>
  summarise(
    products      = n_distinct(product),
    total_revenue = sum(revenue),
    total_units   = sum(units),
    avg_margin    = mean(margin_pct)
  ) |>
  mutate(
    total_revenue = scales::dollar(total_revenue, scale = 1e-3, suffix = "k"),
    total_units   = scales::comma(total_units),
    avg_margin    = scales::percent(avg_margin, accuracy = 0.1)
  ) |>
  rename(Category = category, Products = products, Revenue = total_revenue,
         Units = total_units, `Avg Margin` = avg_margin) |>
  knitr::kable(align = c("l","c","r","r","r"))
```

---

# Key Takeaways

| Package   | What we used it for |
|-----------|---------------------|
| `dplyr`   | `group_by`, `summarise`, `mutate`, `filter`, `arrange` |
| `tidyr`   | `pivot_wider`, `expand_grid` |
| `ggplot2` | bar, line, scatter, facet, heatmap |
| `stringr` | `str_glue`, `str_pad`, `str_detect`, `str_to_upper` |
| `forcats` | `fct_reorder` for sorted factor levels |
| `scales`  | `dollar`, `percent`, `comma` label formatters |

> **Render:** Click **Render** in RStudio, or run `quarto render tidyverse_demo.qmd` in the terminal.
